Efficiently stable presentations from error-correcting codes

### 2023-09-27 efficient_stability.tex
comment in abstract
starter comments

\begin{abstract}
We introduce a notion of \emph{efficient stability} for finite presentations of groups. Informally, a finite presentation using generators $S$ and relations $R$ is \emph{stable} if any map from $S$ to unitaries (say, in some finite von Neumann algebra) that approximately satisfies the relations (in the tracial norm) is close to the restriction of a representation of $G$ to the subset $S$. This notion and variants thereof have been extensively studied in recent years, in part motivated by connections to property testing in computer science. The novelty in our work is the focus on \emph{efficiency} which, informally, places an onus on small presentations (of size, say, logarithmic in the group size) and aims to achieve non-trivial tradeoffs between it and the stability---as, intuitively, more generators and relations place more constraints on the approximate homomorphism, making stability easier to achieve---whereas we wish to minimize the number of variables and constraints.

@@

\paragraph{Motivation.}
A linear error-correcting code $\code$ is a $k$-dimensional subspace of the vector space $\F^n$ over a finite field $\F$ that has certain combinatorial properties. The foremost of these is the \emph{minimal distance} $d$, which is defined as the smallest Hamming weight (number of nonzero coordinates) $|c|$ of a nonzero vector  $c\in\code$. A finer property which concerns us here is the \emph{soundness} of the code, a parameter that is connected to the notion of \emph{local testability}. A code can be (non-uniquely) specified through a \emph{parity-check matrix} $h\in\F^{m\times n}$ as $\code = \ker h$. The rows of $h$ are thought of as constraints (``parity checks'') that uniquely identify elements of $\code$ within $\F^n$. A code is called locally testable \emph{with soundness $\rho$} if for every $x\in \F^n$, $\frac{1}{m}|hx|\geq \rho\, d(x,\code)$, where $d(x,\code)$ denotes the minimum of $|x-c|$ over $c\in \code$. The terminology ``locally testable'' comes from an interpretation of $\frac{1}{m}|hx|$ as the probability of rejection of a natural ``tester'' for $\code$, i.e.\ an algorithm that on input $x$ checks a randomly chosen row $h_i$ of $h$ and accepts if and only if the parity $h_i\cdot x=0$.\footnote{Note that in principle a code can have a small minimal distance $d$, and still be locally testable with soundness $\rho>0$. In general one would like to design families of codes of increasing length $n$, such that $d$ is a linear function of $n$ and $\rho$ is bounded below by a constant independent of $n$.}  Thus a code is called locally testable if words that are far from the code have a high probability of being rejected according to this tester. This notion plays a central role in applications of codes to complexity theory~\cite{babai1991non,PCP_thm}, and continues to be actively studied, see e.g.~\cite{LTC_DELLM,LTC_Panteleev_Kalachev} for a recent breakthrough on the topic.

@@

\paragraph{Group stability.}
This observation, which at first may seem quite anodyne or even forced, opens a universe of questions. The problem of relating approximate representations of a group to exact representations of it is termed \emph{stability} in group theory, and has a long history. There are of course many flavors of the problem, depending on how one defines closeness (should it be the operator norm or the Hilbert-Schmidt norm? Should closeness hold for every relation, or every pair of group elements, or is it sufficient that it holds on average? Etc.) We will review some relevant results in this area below. For now, we can mention Voiculescu's famous counter-example~\cite{voiculescu1983asymptotically} about approximately commuting unitaries, which was motivated by a question of Halmost on pairs of approximately commuting Hermitian operators~\cite{halmos1976some}. Using the terminology of stability, Voiculeschu showed that the presentation $\Z^2 = \langle x,y:[x,y]=e\rangle$ is \emph{not} stable with respect to the operator norm. However, much more recently Glebsky~\cite{glebsky2010almost} showed that the same presentation \emph{is} stable with respect to the normalized Hilbert-Schmidt norm.

This example and many others show that stability is a natural property, which can be very sensitive to how it is measured exactly. Returning to our main concern, so far we have argued that 
stability of approximate \emph{$1$-dimensional} representations of the presentation $G(h)$ is connected to local testability of the code $\ker h$. In the case of $1$-dimensional representations of course the choice of norm does not matter; however, the choice of measuring the error on average over relations, as opposed to e.g.\ taking the maximum, is a first choice that we made and in which we depart from most of the litterature. We will motivate this choice further below; but before we can continue we must pause to introduce the key definitions that our work builds on. For simplicity we restrict the discussion to finite-dimensional representations; in the paper we treat the general case of representations in a tracial von Neumann algebra. For $X\in \C^{d\times d}$ let $\|X\|_f^2 = \frac{1}{d}\Tr(X^* X)$ denote the normalized Hilbert-Schmidt norm.

>>>>

\begin{abstract}
We introduce a notion of \emph{efficient stability} for finite presentations of groups. Informally, a finite presentation using generators $S$ and relations $R$ is \emph{stable} if any map from $S$ to unitaries (say, in some finite von Neumann algebra) that approximately satisfies the relations (in the tracial norm) is close to the restriction of a representation of $G$ to the subset $S$. This notion and variants thereof have been extensively studied in recent years, in part motivated by connections to property testing in computer science. The novelty in our work is the focus on \emph{efficiency} which, informally, places an onus on small presentations (of size, say, logarithmic in the group size) and aims to achieve non-trivial tradeoffs between it and the stability---as, intuitively, more generators and relations place more constraints on the approximate homomorphism, making stability easier to achieve---whereas we wish to minimize the number of variables and constraints. \hnote{This is pretty wordy for an abstract... sometime later this week I can try to shorten it if someone hasn't already done it.}

@@

\paragraph{Motivation.}
A linear error-correcting code $\code$ is a $k$-dimensional subspace of the vector space $\F^n$ over a finite field $\F$ that has certain combinatorial properties. The foremost of these is the \emph{minimal distance} $d$, which is defined as the smallest Hamming weight (number of nonzero coordinates) $|c|$ of a nonzero vector  $c\in\code$. A finer property which concerns us here is the \emph{soundness} of the code, a parameter that is connected to the notion of \emph{local testability}. A code can be (non-uniquely) specified through a \emph{parity-check matrix} $h\in\F^{m\times n}$ as $\code = \ker h$. The rows of $h$ are thought of as constraints (``parity checks'') that uniquely identify elements \hnote{what does it mean for a constraint to ``uniquely'' identify elements of the code?} of $\code$ within $\F^n$. A code is called locally testable \emph{with soundness $\rho$} if for every $x\in \F^n$, $\frac{1}{m}|hx|\geq \rho\, d(x,\code)$, where $d(x,\code)$ denotes the minimum of $|x-c|$ over $c\in \code$. The terminology ``locally testable'' comes from an interpretation of $\frac{1}{m}|hx|$ as the probability of rejection of a natural ``tester'' for $\code$, i.e.\ an algorithm that on input $x$ checks a randomly chosen row $h_i$ of $h$ and accepts if and only if the parity $h_i\cdot x=0$.\footnote{Note that in principle a code can have a small minimal distance $d$, and still be locally testable with soundness $\rho>0$. In general one would like to design families of codes of increasing length $n$, such that $d$ is a linear function of $n$ and $\rho$ is bounded below by a constant independent of $n$.}  Thus a code is called locally testable if words that are far from the code have a high probability of being rejected according to this tester. This notion plays a central role in applications of codes to complexity theory~\cite{babai1991non,PCP_thm}, and continues to be actively studied, see e.g.~\cite{LTC_DELLM,LTC_Panteleev_Kalachev} for a recent breakthrough on the topic.

@@

\paragraph{Group stability.}
This observation, \hnote{I'm not sure we need this parenthetical :)} which at first may seem quite anodyne or even forced, opens a universe of questions. The problem of relating approximate representations of a group to exact representations of it is termed \emph{stability} in group theory, and has a long history. There are of course many flavors of the problem, depending on how one defines closeness (should it be the operator norm or the Hilbert-Schmidt norm? Should closeness hold for every relation, or every pair of group elements, or is it sufficient that it holds on average? Etc.) We will review some relevant results in this area below. For now, we can mention Voiculescu's famous counter-example~\cite{voiculescu1983asymptotically} about approximately commuting unitaries, which was motivated by a question of Halmost on pairs of approximately commuting Hermitian operators~\cite{halmos1976some}. Using the terminology of stability, Voiculeschu showed that the presentation $\Z^2 = \langle x,y:[x,y]=e\rangle$ is \emph{not} stable with respect to the operator norm. However, much more recently Glebsky~\cite{glebsky2010almost} showed that the same presentation \emph{is} stable with respect to the normalized Hilbert-Schmidt norm.

This example and many others show that stability is a natural property, which can be very sensitive to how it is measured exactly. Returning to our main concern, so far we have argued that 
stability of approximate \emph{$1$-dimensional} representations of the presentation $G(h)$ is connected to local testability of the code $\ker h$. In the case of $1$-dimensional representations of course the choice of norm does not matter; however, the choice of measuring the error on average over relations, as opposed to e.g.\ taking the maximum, is a first choice that we made and in which we depart from most of the litterature. We will motivate this choice further below; but before we can continue we must pause to introduce the key definitions that our work builds on. For simplicity we restrict the discussion to finite-dimensional representations; in the paper we treat the general case of representations in a tracial von Neumann algebra. For $X\in \C^{d\times d}$ let $\|X\|_f^2 = \frac{1}{d}\Tr(X^* X)$ denote the normalized Hilbert-Schmidt norm, \hnote{added:} and for a set $S$ let $\mF(S)$ denote the free group generated by the elements of $S$.
